Mapping short reads, faithfully

### 2020-02-10 manuscript.tex
Update materials and methods section
Included a short introduction in the first two sections. Corrected
tiny typos, rewritten some sentences for the sake of clarity. Moved
'MMP implementation' and 'Benchmark' to the Results section, but
its content is not merged there yet.

\subsection{Estimating $N$ from the backward search}

We assume that the true location of the read is a sequence of the genome
that has $N \geq 0$ duplicates. For every nucleotide, the probability that
a duplicate differs from the target is a constant $\mu$, and the
probability that the read differs from the target is a constant $p$. We
further assume that the genome has size $G$ and that its composition is a
equiprobable random mix of the four nucleotides.

@@

\subsubsection{Super quality} Super reads are defined as reads that
satisfy the two following conditions: 20-mers extracted every 10
nucleotides all have a single hit in the genome (see the Results section
and Figure~\ref{fig_supertest}), and the read is mapped without mismatch.

Super reads are mapped to an incorrect location if $i.$ the target has
exactly one duplicate, $ii.$ the read contains $m$ errors, $iii.$ the
paralog differs from the target exactly at the $m$ locations of the
errors, $iv.$ the $m$ incorrect nucleotides match the paralog, and $v.$
the $m$ errors are in the odd-numbered 10-mers of the read. The
justification for these conditions are explained in more detail in
Figure~\ref{fig_supread}.

Condition $v.$ imposes $m = 1 + \lfloor L/20 \rfloor$ for a read of size
$L$. The probability of $i.$ is denoted $\alpha$. The probability of $ii.$
is ${L \choose m}p^m(1-p)^{L-m}$, where $p$ is the error rate of the
sequencer. The probability of $iii.$ is $\mu^m(1-\mu)^{L-m}$, where $\mu$
is the probability that the duplicate differs from the target. The
probability of $iv.$ is $1/3^m$. The probability of $v.$ is $10^m/{L
\choose m}$. We need to multiply those terms together and divide them by
the probability that a read is super, which is fraction $\beta$ of the
probability that a read has no error. The mapping quality of super reads
is finally computed as
\begin{align}
\label{eq_super}
\frac{\alpha}{\beta} \left( \frac{3.33 \cdot p\mu}{1-p}
\right)^m (1-u)^{L-m}.
\end{align}

@@

It was verified empirically, that expression (\ref{eq_bayes}) is an
overestimate when $x = 0$. In this case, the estimate is replaced by a
direct approach where we compute the probability that a paralog matches
the read perfectly. More specifically, we compute the probability that the
read has exactly one error and that a paralog matches the incorrect
sequence. In this case, the probability that the location is incorrect is
computed as
\begin{align}
\label{eq_x0}
P(N \geq 1) \cdot \frac{Lp(1-p)^{L-1}
\left(1-\left((1-\mu)^{L-1}\mu/3\right)^N\right)}{(1-p)^L}.
\end{align}

@@

\subsubsection{Low quality} Reads with initial estimate $N > 20$ are
mapped by checking only the longest MEM seed. A seed of size $S$ exists if
there are $S$ nucleotides without error or if there are $S$ nucleotides
with an error matching at least one of the $N$ paralogs. In the first
case, the target is discovered, in the second it is not. We also have to
count whether seed is limited by errors or by the end of the read.
The probability of the first case is
\begin{align*}
P_{\text{hit}} = \delta_L \cdot (1-p)^S \cdot \delta_R,
\end{align*}
where $\delta_{L/R} = 1$ if the seed is flanked by the end of the read and
$p$ otherwise.

Counting only the most frequent case with a single error, the probability
of the second case is
\begin{equation*}
\begin{split}
P_{\text{miss}} = \; &\delta_L \cdot Sp(1-p)^{S-1} \cdot \delta_R  \\
   &\times \left(1- \left( 1-\delta_L' \cdot(1-\lambda)^{S-1}\lambda/3
   \cdot \delta_R'\right)^N \right),
\end{split}
\end{equation*}
where $\lambda$ is defined as above, and where $\delta_{L/R}' = 1$ if the
seed is flanked by the end of the read and $\lambda$ otherwise. Applying
Bayes' formula, the mapping quality is finally computed as

@@

\subsubsection{Estimating $P(N \geq 1)$.} This quantity appears in
expressions (\ref{eq_bayes}) and (\ref{eq_x0}). The estimate is based on
the method shown in Figure~\ref{fig_back}, where we use the end of the
read with the shortest query. We consider three cases: $i.$ the target has
a paralog, $ii.$ only one end of the target has a paralog, and $iii.$ the
target is unique. Case $ii.$ can occur when the read straddles the end of
a transposon or other repeated sequence. We consider that $N \geq 1$ only
in case $i.$ where the sequence has a full-length paralog.

>>>>

\subsection{Estimating the number of duplicates}

The main source of uncertainty in the measurements of mapping quality
estimation is the existence of sequence duplicates. Duplicates are groups
of sequences that are not exactly identical but share high levels of
similarity, likely because they co-evolved together or resulted from
horizontal DNA transfer events. Duplicate sequences are hard to map
reliably because the similarities between the read and the secondary
candidate references are not negligible. At the same time, the exact
probability that the read originated from one of the duplicates is
difficult to quantify in practice. In this section we introduce a method
to estimate the number of duplicates of a sequence. This estimate is
crucial in the whole process of mapping quality computation because the
number of duplicates will be the main determinant of the quality range.

We assume that the true location of the read is a sequence of the genome
that has $N \geq 0$ duplicates. For every nucleotide, the probability that
a duplicate differs from the target is a constant $\mu$, and the
probability that the read differs from the target is a constant $p$. We
further assume that the genome has size $G$ and that its composition is an
equiprobable random mix of the four nucleotides.

@@

Based on the estimate of the number of duplicates, we define three
different groups of reads. The mapping quality for each read group
is computed differently because the dominant mapping uncertainty
depends on the number of duplicates.

\subsubsection{Super quality} Super reads are defined as reads that
satisfy the two following conditions: 20-mers extracted every 10
nucleotides all have a single hit in the genome (see the Results section
and Figure~\ref{fig_supertest}), and the read is mapped without mismatch.
Super reads are mapped to an incorrect location iif:
\begin{enumerate}
  \item the read contains $m$ errors
  \item the target has exactly one duplicate
  \item the duplicate differs from the target exactly at the $m$ locations
    of the errors
  \item the $m$ incorrect nucleotides match the sequence of the duplicate
  \item the $m$ errors are in the odd-numbered 10-mers of the read
\end{enumerate}
The justification for these conditions are explained in more detail in
Figure~\ref{fig_supread}.

The last condition imposes $m = 1 + \lfloor L/20 \rfloor$ for a read of size
$L$. The probability of condition $1$ is denoted $\alpha$. The probability
of condition $2$ is ${L \choose m}p^m(1-p)^{L-m}$, where $p$ is the error
rate of the read sequencer. The probability of condition $3$ is
$\mu^m(1-\mu)^{L-m}$, where $\mu$ is the probability that the duplicate
differs from the target. The probability of condition $4$ is $1/3^m$ and the
probability of condition $5$ is $10^m/{L\choose m}$.

We need to multiply the probability of each condition together and divide
them by the probability that a read is a \emph{super read}, which is fraction
$\beta$ of the probability that a read has no error. The mapping quality of
super reads is finally computed as
\begin{align}
\label{eq_super}
\frac{\alpha}{\beta} \left( \frac{3.33 \cdot p\mu}{1-p}
\right)^m (1-u)^{L-m}.
\end{align}

@@

It was verified empirically that expression (\ref{eq_bayes}) is an
overestimate when $x = 0$. In this case, the estimate is replaced by a
direct approach where we compute the probability that a paralog matches
the read perfectly. More specifically, we compute the probability that the
read has exactly one error and that a paralog matches the incorrect
sequence. In this case, the probability that the location is incorrect is
computed as
\begin{align}
\label{eq_x0}
P(N \geq 1) \cdot \frac{Lp(1-p)^{L-1}
\left(1-\left((1-\mu)^{L-1}\mu/3\right)^N\right)}{(1-p)^L}.
\end{align}

@@

\subsubsection{Low quality} Reads with initial estimate $N > 20$ are
mapped by checking only the longest MEM seed. A seed of size $S$ exists if
there are $S$ nucleotides without error or if there are $S$ nucleotides
with possibly many errors matching at least one of the $N$ paralogs. In the
first case, the target is discovered, in the second it is not. We also have
to account whether the seed is flanked by errors or by the end of the read.
The probability of the first case is
\begin{align*}
P_{\text{hit}} = \delta_L \cdot (1-p)^S \cdot \delta_R,
\end{align*}
where $\delta_{L/R} = 1$ if the seed is flanked by the end of the read and
$p$ otherwise.

Considering only the most frequent case with a single error, the probability
of the second case is
\begin{equation*}
\begin{split}
P_{\text{miss}} = \; &\delta_L \cdot Sp(1-p)^{S-1} \cdot \delta_R  \\
   &\times \left(1- \left( 1-\delta_L' \cdot(1-\lambda)^{S-1}\lambda/3
   \cdot \delta_R'\right)^N \right),
\end{split}
\end{equation*}
where $\lambda$ is defined as above, and where $\delta_{L/R}' = 1$ if the
seed is flanked by the end of the read and $\delta_{L/R}' = \lambda$ otherwise.
Applying Bayes' formula, the mapping quality is finally computed as

@@

\subsubsection{Estimating $P(N \geq 1)$} This quantity appears in
expressions (\ref{eq_bayes}) and (\ref{eq_x0}). The estimate is based on
the method shown in Figure~\ref{fig_back}, where we use the end of the
read with the shortest query. We consider three cases: $i.$ the target has
a paralog, $ii.$ only one end of the target has a paralog, and $iii.$ the
target is unique. Case $ii.$ can occur when the read straddles the end of
a transposon or other repeated sequence. We consider that $N \geq 1$ only
in case $i.$ where the sequence has a full-length paralog.
